Supp mat: time series paper

### 2017-05-13 writing/manuscript_draft.Rmd
added to discussion section

<!-- 
- Findings and how they relate to past work
- Extensions to other systems. Useful to use similar species
- Biological variables not good predictors, but we also used all long-lived vertebrate species
- Other questions to ask and results in appendix
- Future recommendations to others and future work
- implications
- conclusion paragraph

@@

There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...

>>>>

<!-- 
- Findings and how they relate to past work
- Extensions to other systems. Useful to use similar species
- Biological variables not good predictors, but we also used all long-lived vertebrate species
- Other questions to ask and results in appendix

- Future recommendations to others and future work
- implications
- conclusion paragraph

@@

We also found that $t_{min}$ did not correlate with any biological variables of interest (cite figure or supp mat). We initially hypothesized that species with longer lifespans or generation times may require more years to sample. (Our result could have been the effect of two reasons) First, the database we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreases, is only worried about population surveys. Therefore, biological variables may be more important if the question itself was different...

### Other questions addressed

In the appendix we asked two additional questions to make estimates of $t_{min}$. First, we asked how many years of sampling are neccessary to estimate a geometric growth rate?

### Past work on this issue
There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...

An important related question, is the optimal allocation of sampling effort in space versus time (citations). In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (cite figure here). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a longterm census of the collared pika (\emph{Ochotona collaris}) found in the Yukon. They estimated longterm growth rates among three subpopulations over a 10-year period. They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby, may be difficult. Therefore, they argued more sampling should be conducted spatially than temporally [@Morrison2008].

### 2017-05-13 writing/manuscript_draft.tex
added to discussion section

\begin{itemize}
\tightlist
\item
  Findings and how they relate to past work
\item
  Extensions to other systems. Useful to use similar species
\item
  Biological variables not good predictors, but we also used all
  long-lived vertebrate species
\item
  Other questions to ask and results in appendix
\item
  Future recommendations to others and future work
\item
  implications
\item
  conclusion paragraph
\end{itemize}

The IUCN has an 10 year/3 generation rule\ldots{} If we apply this rule
to the species in our database we see that approximately 50.5760369\% of
populations need a longer minimum time series than that required by
IUCN.

>>>>

We show two different approaches to estimate the minimum time series
length required required to address a particular question of interest.
The simulation-based approach has been suggested by others, especially
in situations where analysis is more complicated than a simple power
analysis (Gerrodette 1987). We focus on a empirical approach to estimate
the minimum time series length requried.

The first question we asked was, what is \(t_{min}\) for determining
whether a population was increasing or decreasing over time according to
linear regression. We found that for 868 populations, at least 10-15
years of continuous monitoring were neccessary. In line with theoretical
predictions (citations), we also found \(t_{min}\) was strongly
correlated with the trend strength, variability in population size, and
temporal autocorrelation.

We also found that \(t_{min}\) did not correlate with any biological
variables of interest (cite figure or supp mat). We initially
hypothesized that species with longer lifespans or generation times may
require more years to sample. (Our result could have been the effect of
two reasons) First, the database we used may not include a diverse
enough set of species with different life history traits. Second, the
question we poised, whether a population is increasing or decreases, is
only worried about population surveys. Therefore, biological variables
may be more important if the question itself was different\ldots{}

\subsubsection{Other questions
addressed}\label{other-questions-addressed}

In the appendix we asked two additional questions to make estimates of
\(t_{min}\). First, we asked how many years of sampling are neccessary
to estimate a geometric growth rate?

\subsubsection{Past work on this issue}\label{past-work-on-this-issue}

There have been previous attempts to address how long a time series
should be to assess trends in abundance. For example,\ldots{}

An important related question, is the optimal allocation of sampling
effort in space versus time (citations). In a theoretical investigation
of this question, Rhodes and Jonzen (2011) found that the optimal
allocation of sampling depended strongly on temporal and spatial
autocorrelation. If spatial population dynamics were highly correlated,
then it was better to sample more temporally, and vice versa. Our work
supports this idea as populations with strong temporal autocorrelation
needed less years of sampling (cite figure here). Morrison and Hik
(2008) also studied the optimal allocation of sampling effort in space
versus time, but used emprical data from a longterm census of the
collared pika (\emph{Ochotona collaris}) found in the Yukon. They
estimated longterm growth rates among three subpopulations over a
10-year period. They found that censuses less than 5 years may be
misleading and that extrapolating from one population to another, even
when nearby, may be difficult. Therefore, they argued more sampling
should be conducted spatially than temporally (Morrison and Hik 2008).

\subsubsection{Limitations}\label{limitations}

Our work has obvious limitations in determining the minimum time series
length required. First, \(t_{min}\) depends strongly on the question
being asked and the availability of long-term census data. For our
empirical approach, the subsampling of the full time series allows for
estimates of power, but the individual subsamples are clearly not
independent of one another. (sentence about alpha and power
limititations here) The perfect scenario would be to have a population
model built and parameterized for each population of interest. Then,
model simulations could be used to estimate the minimum time series
reuqired to address a specific question of interest. Clearly, this is
not always practical, especially if running analyses for a wide array of
species as we do here. In addition, our statistical models suggest that
\(t_{min}\) does not correlate with any biological variables of
interest, at least for the question of linear regression. Therefore, it
is not possible to use these results to predict \(t_{min}\) for another
population, even if the population is of a species with a similar
life-history to one in our database.

\subsubsection{Conclusions}\label{conclusions}

\begin{itemize}
\tightlist
\item
  what we found, implications, future work
\end{itemize}

We use a database of 868 populations to determine the minimum time
series length required. We show that to assess longterm trends in
abundance, 10-15 years of continuos monitoring are required. In line
with theoretical predictions, we also show that \(t_{min}\) is strongly
correlated with the overall trend in abundance, variability in
abundance, and the temporal autocorrelation. Our work implies that for
many populations, census data less than 10-15 years are probably not
reliable. This stresses the importance of longterm monitoring programs.
(where do IUCN recommendations come from?) From both a scientific and
management perspective estimates of \(t_{min}\) are important. If a time
series is too short, we are likely to make incorrect conclusions about
longterm trends. In addition, a time series that is too long may have
been a poor use of scare funds. Future work can examine other species,
with a wider range of life history characteristics, or address other
questions besides longterm trends in abundance.
